---
title: <span style="color:red">A simple analysis of dataset '**mtcars**'</span>
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(rmarkdown)
library(tibble)
library(DT)
df = mtcars
```

## **Description of dataset [mtcars](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html)**
The dataset '**mtcars**' contains 11 numeric variables with a total of 32 observations.The variables are: 
<br>
<br>
<font size="4">1. **mpg**  -	Miles/(US) gallon</font>
<br>
<font size="4">2. **cyl**  - Number of cylinders</font>
<br>
<font size="4">3. **disp** - Displacement (cu.in.)</font>
<br>
<font size="4">4. **hp**   - Gross horsepower</font>
<br>
<font size="4">5. **drat** - Rear axle ratio</font>
<br>
<font size="4">6. **wt**   - Weight (1000 lbs)</font>
<br>
<font size="4">7. **qsec** - 1/4 mile time</font>
<br>
<font size="4">8. **vs**   - Engine (0 = V-shaped, 1 = straight)</font>
<br>
<font size="4">9. **am**   - Transmission (0 = automatic, 1 = manual)</font>
<br>
<font size="4">10. **gear** - Number of forward gears</font>
<br>
<font size="4">11. **carb** - Number of carburetors</font>                   

## Pre-processing data

### **Checking for missing values**

```{r mtcars, echo = TRUE}
anyNA(mtcars)
```

<font size="4">The **'anyNA'** function is used to check if the dataset contains any missing values. As the ouput of the function is **FALSE**, it indicates all non-missing values in the dataset.</font>

## Pre-processing data

<font size="5">**Checking for skewness-**</font>
<br>
<font size="4">The variables **cyl, disp, drat, wt, qsec** are normally distributed. For example, see below diagram of variables **drat** (left) and **qsec** (right).</font>  
```{r}
library(rcompanion)
attach(mtcars)
opar <- par(no.readonly=TRUE)
par(mfrow=c(1,2))
plotNormalHistogram(mtcars$drat)
plotNormalHistogram(mtcars$qsec)
par(opar)
detach(mtcars)
```

## Pre-processing data

<font size="4">The varibales **hp, mpg, gear** and **carb** are skewed towards the right. For example, see below diagram of varibales **hp** (left) and **mpg** (right).</font> 

```{r}
library(rcompanion)
attach(mtcars)
opar <- par(no.readonly=TRUE)
par(mfrow=c(1,2))
plotNormalHistogram(mtcars$hp)
plotNormalHistogram(mtcars$mpg)
par(opar)
detach(mtcars)
```

## **Scatterplots**

<font size="4">Variables **wt** and **mpg** (left) is negatively related as points in the graph show a downhill pattern as we move from left to right while variables **wt** and **disp** (right) are positively related to each other as the data points for x increase as y values increase.</font> 
```{r}
attach(mtcars)
opar <- par(no.readonly=TRUE)
par(mfrow=c(1,2))
plot(wt,mpg, main="Scatterplot of wt vs. mpg")
plot(wt,disp, main="Scatterplot of wt vs. disp")
par(opar)
detach(mtcars)
```


## **Linear Regression**
<font size="4">We use a simple linear regression function to predict miles/gallon (**mpg** - response variable) travelled from weight (**wt** - predictor variable) of the car (shown below).</font>
```{r, echo = TRUE}
fit <- lm(mpg ~ wt, data=mtcars)
```
<font size="4">The summary of the function is presented in the next slide. We can see from the output that the model fits our data moderately well as indicated by the **Multiple R-squared value** of **0.75**.  This indicates that the model accounts for **75%** of the variance in mpg (miles/gallon).</font> 
<br>
<br>
<font size="4">We can also see that the **regression coefficient (-5.3)** is significantly different from zero
<br>
(p < 0.001) and indicates that there’s an expected decrease of 5.3 units of mpg for every 1 unit increase in wt.</font>

## Summary output of Regression model
```{r} 
summary(fit)
```


## Regression Plot
<font size="4">The plot shows the regression line fitted by the model. As the data values lie around this line, this indicates that the model fits our data well.</font> 

```{r}
plot(mtcars$wt, mtcars$mpg, xlab="wt",ylab="mpg", main = "Regression of Miles/Gallon (mpg) on Weight (wt)" )
abline(fit)
```

