---
title: "An Simple Analysis of Dataset 'mtcars'"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(rmarkdown)
library(tibble)
library(DT)
df = mtcars
```

>##### This R Markdown document has been created to explore the dataset [mtcars](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html) located in the R Datasets Package for [Assignment 1](https://stat545.stat.ubc.ca/evaluation/hw01/hw01/)


#### **Description of dataset 'mtcars'**
>
>##### The datatset **'mtcars'** is a dataset that contains 11 numeric variables with a total of 32 observations. Full description of the dataset can be found [here](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).
>

#### **The variables are:-**
>
| Variable | Description |
|----------|-------------|
| mpg      | Miles/(US) gallon |
| cyl      | Number of cylinders |                     
| disp     | Displacement (cu.in.) |                   
| hp       | Gross horsepower |                        
| drat     | Rear axle ratio |                         
| wt       | Weight (1000 lbs) |                       
| qsec     | 1/4 mile time |
| vs       | Engine (0 = V-shaped, 1 = straight) |
| am       | Transmission (0 = automatic, 1 = manual) |
| gear     | Number of forward gears |                 
| carb     | Number of carburetors |   
>

#### **Dataset Structure**
>##### The **structure** of the dataset can be seen below-
>
```{r mtcars}
str(mtcars)
```
>
>##### The variables **vs** (engine) and **am** (Transmission) are factor variables with two levels (0 and 1).


#### **Checking for missing values**
>##### The **'anyNA'** function is used to check if the dataset contains any missing values. As the ouput of the function is **FALSE**, it indicates all non-missing values in the dataset.

```{r}
anyNA(mtcars)
```



#### **Checking for Skewness**
>##### The variables **cyl, disp, drat, wt, qsec** are **normally distributed**. (Graphs in order of the variables listed).  

```{r}
library(rcompanion)
attach(mtcars)
opar <- par(no.readonly=TRUE)
par(mfrow=c(3,2))
plotNormalHistogram(mtcars$cyl)
plotNormalHistogram(mtcars$disp)
plotNormalHistogram(mtcars$drat)
plotNormalHistogram(mtcars$wt)
plotNormalHistogram(mtcars$qsec)
par(opar)
detach(mtcars)
```

>##### The varibales **hp, mpg, gear** and **carb** are **right skewed**. (Graphs in order of the variables listed).

```{r}
library(rcompanion)
attach(mtcars)
opar <- par(no.readonly=TRUE)
par(mfrow=c(2,2))
plotNormalHistogram(mtcars$hp)
plotNormalHistogram(mtcars$mpg)
plotNormalHistogram(mtcars$gear)
plotNormalHistogram(mtcars$carb)
par(opar)
detach(mtcars)
```



#### **Scatterplots (to identify relationship between variables)**
>
>##### Below are the sample scatterplots of variables **wt vs mpg** and **wt vs disp**. The variables **wt** and **mpg** is negatively related as the points in the graph show a downhill pattern as we move from left to right while the variables **wt** and **disp** are positively related to each other as the data points for x increase as y values increase. 

```{r}
attach(mtcars)
opar <- par(no.readonly=TRUE)
par(mfrow=c(1,2))
plot(wt,mpg, main="Scatterplot of wt vs. mpg")
plot(wt,disp, main="Scatterplot of wt vs. disp")
par(opar)
detach(mtcars)
```



#### **Linear Regression**
>
>##### We use a simple linear regression function to predict miles/gallon (**mpg** - response variable) travelled from weight (**wt** - predictor variable) of the car (shown below). The analysis is provided in the listing below and the resulting graph is presented.
>

```{r}
fit <- lm(mpg ~ wt, data=mtcars)
summary(fit)
```

>##### We can see from the output that the model fits our data moderately well as indicated by the **Multiple R-squared value** of **0.75**.  This indicates that the model accounts for **75%** of the variance in mpg (miles/gallon).
>
>##### We can also see that the **regression coefficient (-5.3)** is significantly different from zero (p < 0.001) and indicates that there’s an expected decrease of 5.3 units of mpg for every 1 unit increase in wt.

#### **Regression plot**

```{r}

plot(mtcars$wt, mtcars$mpg, xlab="wt",ylab="mpg", main = "Regression of Miles per Gallon (mpg) on Weight (wt)" )
abline(fit)
```

>
>##### The plot shows the regression line fitted by the model. As the data values lie around this line, this indicates that the model fits the data well.
>






